---
title: "Chi-Squared Residuals"
author: "davegoblue"
date: "March 11, 2016"
output: html_document
---

This document is to validate how R calculates chi-squared, Cramer's V, and standardized residuals for a simple test of categorical association.  The test data frame will be a small 3x3 example.

```{r}
testFrame <- data.frame(colI = c(5, 34, 33), colII = c(6, 47, 32), 
                        colIII = c(9, 48, 14), row.names=c("A","B","C")
                        )
testFrame
```

Next, expected values are calculated for each cell as N * Prow * Pcol.  Further, residuals are calculated as (observed - expected):

```{r}
expMatrix <- as.matrix(rowSums(testFrame)) %*% t(as.matrix(colSums(testFrame)))
expMatrix <- expMatrix / sum(testFrame)
round(expMatrix,1)

resMatrix <- testFrame - expMatrix
round(resMatrix,1)
```

Then, we run (by hand) each of the chi-squared, df, Cramer's V, and standardized residuals:  

```{r}
testChiSq <- sum(resMatrix^2 / expMatrix)
nR <- nrow(resMatrix)
nC <- ncol(resMatrix)

dfChiSq <- (nR - 1) * (nC - 1)
cramerV <- sqrt(testChiSq / (nR * nC) / (min(nR, nC) - 1) )

## Pearson residuals - (obs - exp) / sqrt(exp)
prsMatrix <- resMatrix / sqrt(expMatrix)

## Standardized residuals - (obs - exp) / sqrt(V)
## Vij = sqrt(Expij * (1 - pRow) * (1 - pCol))
stdMatrix <- resMatrix / 
             sqrt(expMatrix * 
                   (as.matrix(1 - rowSums(expMatrix)/sum(expMatrix)) 
                    %*%             
                    t(as.matrix(1-colSums(expMatrix)/sum(expMatrix)))
                    )
                  )

print(paste0("Chi-squared is ",round(testChiSq,2)," with df=",dfChiSq,
             " (p=",
             round(pchisq(testChiSq, df=dfChiSq, lower.tail=FALSE),4),
             ")"
             )
      )

print(paste0("Cramer's V is: ",round(cramerV,3)))

## Pearson residuals - (obs - exp) / sqrt(exp)
round(prsMatrix,2)

## Standardized residuals - (obs - exp) / sqrt(V)
## Vij = sqrt(Expij * (1 - pRow) * (1 - pCol))
round(stdMatrix,2)
```

The results are compared with just running the chi-squared test in R, listing each of the 9 dimensions in order.  Note that following as per residual calculations in chisq.test:  
  
* "residuals"" are the Pearson residuals (observed - expected) / sqrt(expected)  
* "stdres" are the standardized residuals (observed - expected) / sqrt(V)  
* Vij = sqrt[ EXPij x (1 - P_i_ALL) x (1 - P_ALL_j ) ]  
  
```{r}
testChi <- chisq.test(testFrame)
for (intCtr in 1:length(testChi)) {
    print(testChi[intCtr])
}

## Borrowed from http://www.r-bloggers.com/example-8-39-calculating-cramers-v/
## With adaptations
cv.test <- function(x) {
    CV <- sqrt(chisq.test(x, correct=FALSE)$statistic /
               (sum(!is.na(x)) * (min(nrow(x), ncol(x)) - 1) )
               )
  
    print.noquote("Cramér V / Phi:")
  
    return(as.numeric(CV))
}

cv.test(testFrame)
```

The hand-calculated results match up with the R pchisq.test outputs in this particular example.  