---
title: "Inferential Stats - Condensed Chapter Notes"
author: "davegoblue"
date: "March 13, 2016"
output: html_document
---

# Synopsis and Overview  
This document contains some of the key concepts, statistics, and R formulae from the UvA Inferential Statistics course.  It is intended as a helpful reference sheet for future use.  

## Module 1: Comparing Two Groups  
There are two schools of thought, frequentist (more common) and Bayesian (growing):  
  
* Frequentist - what are the odds of observing X given my hypothesized population?  
* Bayesian - what are the odds of a hypothesized population given that I observed X?  

There are a few good cautions raised about significance and confidence levels:  
  
* Statistical Significance - The x% confidence interval is constructed such that if we run the experiment infinite times, x% of our intervals would contain the true population mean  
* Practical Significance - does it actually matter that Ho: P=0 is rejected?  The confidence interval helps to assess effect magnitude  

Recall that there are two key error types:  
  
* Type I (reject a true null) - happens (1 - alpha), where alpha is pre-set  
* Type II (fail to reject a false null) -- happend beta, so (1 - beta) is the test power  
* Power is increased by better instruments, more homogenous samples, larger N, larger effect, one-sided tests, parametric tests (for sufficiently large N)  

####_Two indepedent proportions_  
The main metric is a z-test, and it generally requires at least 10 positive and 10 negative per group for one-sided tests, and at least 5 positive and 5 negative per group for two-sided tests:  

The test statistic is designed to follow a z-distribution, specifically:  
  
* Since null is equality, the test stastic starts as (p1-hat - p2-hat)  
* Standard error is then sqrt(p-hat * (1 - p-hat) * (1/n1 + 1/n2))  
* P-hat is the pooled proportion, or (n1 * p1-hat + n2 * p2-hat) / (n1 + n2)  
* CI (after the fact) can no longer use the pooled proportion since null is no longer assumed  
* (p1-hat - p2-hat) +/- z(alpha/2) * sqrt(p-hat1 * (1-p-hat1) / n1 + p-hat2 * (1-p-hat2) / n2)  

This can also be expressed as a relative risk, or p1/p2, with its own assumed confidence intervals.  

See below for an example from two random normal distributions:  

```{r}
set.seed(0313160758)
norm1 <- rnorm(100,mean=2,sd=1)
norm2 <- rnorm(120,mean=2.5,sd=1.5)
p1Norm <- norm1 > 2
p2Norm <- norm2 > 2
p1Hat <- mean(p1Norm)
p2Hat <- mean(p2Norm)
n1 <- length(p1Norm)
n2 <- length(p2Norm)

## Calculate means, difference, pooled proportion, standard error
print(paste0("Means of the data are P1: ",p1Hat," and P2: ",p2Hat))
poolProp <- (n1*p1Hat + n2*p2Hat) / (n1 + n2)
stdError <- sqrt(poolProp * (1-poolProp) * (1/n1 + 1/n2))
print(paste0("Pooled proportion is: ",round(poolProp,3)," with stderr: ",round(stdError,3)))

## Calculate test statistic and p-value
zTestStat <- (p1Hat - p2Hat) / stdError
pTwoSided <- 1 - 2 * abs((pnorm(zTestStat) - 0.5))
print(paste0("The z-stat of: ",round(zTestStat,3)," which has z^2: ",round(zTestStat^2,3),
             " has two-sided significance: ",round(pTwoSided,3)
             )
      )

## Calculate post-hoc CI
newStdErr <- sqrt(p1Hat * (1 - p1Hat) / n1 + p2Hat * (1 - p2Hat) / n2)
critZ <- qnorm(.05/2, lower.tail=FALSE)
print(paste0("Post-hoc 95% CI for difference in proportions is ",
             paste(round(p1Hat - p2Hat + c(-1,1) * critZ * newStdErr, 3), collapse=" , ")
             )
      )

## Comparison to R -- note that R uses chi-squared which is directly related to N(0.1)
prop.test(x=c(sum(p1Norm),sum(p2Norm)) , n=c(n1, n2) , correct=FALSE)

```
  
And, we observe that the findings from R match the hand calculations, with the exception that R runs a chi-squared test and the hand calculations are a z-test.  If x~N(0,1) then x^2 ~ chi-squared df=1, so these are functionally the same approach and outcome.  
  
####_Two indepedent means_  
For two indepednent means, the t-test is applied.  This requires independence of samples, roughly normal population distributions (though t is robust against this), and a "large enough" (N ~30) sample:  
  
The test statistic is designed to follow a t-distribution, specifically:  
  
#####_Unequal variance_  
* Since null is equality, the test stastic starts as (x1-hat - x2-hat)  
* Pooled standard error is then unPSE = sqrt(s1^2/n1 + s2^2/n2)  
* df is a cluster - (s1^2/n1 + s2^2/n2)^2 / [ (s1^2/n1)^2 / (n1 - 1) + (s2^2/n2)^2 / (n2 - 1) ]  
* Test-statistic is (x1-hat - x2-hat) / unPSE, following t with df  
* Post-hoc CI is (x1-bar - x2-bar) +/- t(alpha/2) * unPSE 
  
#####_Assume equal variance_  

* Pooled standard error is then eqPSE = sqrt( [ (n1-1) * s1^2 + (n2-1) * s2^2 ] / [ (n1 - 1) + (n2 - 1) ]  
* df is easy - n1 + n2 - 2    
* Test-statistic is (x1-hat - x2-hat) / eqPSE, following t with df  
  
Note that the unequal variances approach is greatly preferred, as the equal variances approach can be sensitive to violations of normality.  

See below for an example from two random normal distributions (we will keep the norm1 and norm2 as originally drawn, but without any conversions for "greater than 2":  

```{r}
x1Hat <- mean(norm1)
x2Hat <- mean(norm2)
s1 <- sd(norm1)
s2 <- sd(norm2)
n1 <- length(p1Norm)
n2 <- length(p2Norm)

## Report on individual means, standard deviations, and sample sizes
print(paste0("Distribution 1 has mean: ",round(x1Hat,2)," with std ",round(s1,2)," on n=",n1))
print(paste0("Distribution 2 has mean: ",round(x2Hat,2)," with std ",round(s2,2)," on n=",n2))

## Run this as an unequal variance approach
unPSE <- sqrt(s1^2/n1 + s2^2/n2)
unDF <- (s1^2/n1 + s2^2/n2)^2 / ((s1^2/n1)^2 / (n1-1) + (s2^2/n2)^2 / (n2-1))
print(paste0("Assuming unequal variance, we have difference: ",round(x1Hat - x2Hat,2),
             " with pooled SE: ",round(unPSE,2)," and df = ",round(unDF,1)
             )
      )

unTest <- (x1Hat - x2Hat) / unPSE
unPTwoSided <- 1 - 2 * abs(pt(unTest,df=unDF) - 0.5)
print(paste0("The t-statistic ",round(unTest,3)," has two-sided significance ",round(unPTwoSided,3)))

critT <- qt(.05/2,df=unDF,lower.tail=FALSE)
print(paste0("The 95% CI for difference in means is ",
             paste(round(x1Hat - x2Hat + c(-1,1) * critT * unPSE, 3), collapse=" , ")
             )
      )

## Comparison to the R results
t.test(norm1,norm2,paired=FALSE,var.equal=FALSE)


## Run this as an equal variance approach
eqS <- sqrt( ( (n1-1) * s1^2 + (n2-1) * s2^2 ) / ( (n1 - 1) + (n2 - 1) ) )
eqPSE <- eqS * sqrt(1/n1 + 1/n2)
eqDF <- n1 + n2 - 2
print(paste0("Assuming equal variance, we have difference: ",round(x1Hat - x2Hat,2),
             " with S: ",round(eqS,2)," and pooled SE: ",round(eqPSE,2)," with df = ",round(eqDF,1)
             )
      )

eqTest <- (x1Hat - x2Hat) / eqPSE
eqPTwoSided <- 1 - 2 * abs(pt(eqTest,df=eqDF) - 0.5)
print(paste0("The t-statistic ",round(eqTest,3)," has two-sided significance ",round(eqPTwoSided,3)))

critT <- qt(.05/2,df=eqDF,lower.tail=FALSE)
print(paste0("The 95% CI for difference in means is ",
             paste(round(x1Hat - x2Hat + c(-1,1) * critT * eqPSE, 3), collapse=" , ")
             )
      )

## Comparison to the R results
t.test(norm1,norm2,paired=FALSE,var.equal=TRUE)
```

And, we observe that the findings from R match the hand calculations.  
  
####_Two dependent proportions_  
With two dependent proportions, you have a McNemar test.  Note that R will run this as a chi-squared (worry about the direction for one-sided in interpretation) while UvA runs this as a z-test (worry about the direction in the statistic set-up).  
  
Essentially, there are four things that can happen:  Good -> Good, Good -> Bad, Bad -> Good, and Bad -> Bad.  We are only actually interested in Good -> Bad (GB) and Bad -> Good (BG).  

* z = (BG - GB) / sqrt(BG + GB)  
* chiSq = (BG - GB)^2 / (BG + GB)  

Suppose we have a grid of 311, 34 (BG), 17 (GB), 14.  We only care about the 34 and the 17:  

```{r}
zTest <- (34 - 17) / sqrt(34 + 17)
zPStat <- 1 - 2 * abs(pnorm(zTest) - 0.5)
print(paste0("Using z-test, we have z: ",round(zTest,3)," with two-sided p: ",round(zPStat,3)))

## Compare to R
mcnemar.test(x=matrix(data=c(311,34,17,14),nrow=2,byrow=TRUE),correct=FALSE)
```

Except for the use of chi-squared vs. z-statistic, the outputs are substantially the same.  
  
####_Two dependent means_  
This test is much simpler to run since you have paired observations.  Essentially:  
* Test-statistic t = (x1-bar - x2-bar) / ( Sd / sqrt(n) )  
* Sd is the standard deviation of the differences  
* df = n-1  
* CI = (x1-bar - x2-bar) +/- t(alpha/2) * Sd / sqrt(n)  

Below is an example for a random normal added to a random normal:  

```{r}
myVar1 <- norm1
myVar2 <- norm1 + rnorm(100,mean=0.5,sd=5)
myDiff <- myVar1 - myVar2

muDiff <- mean(myDiff)
sDiff <- sd(myDiff)
nDiff <- length(myDiff)
seDiff <- sDiff / sqrt(nDiff)
dfDiff <- nDiff - 1

tStatDiff <- muDiff / seDiff
pStatDiff <- 1 - 2 * abs(pnorm(tStatDiff) - 0.5)

print(paste0("The difference in means is: ",round(muDiff,2)," with standard error: ",round(seDiff,2)))
print(paste0("The t-statistic ",round(tStatDiff,2)," with df = ",dfDiff,
             " has two-sided significance ",round(pStatDiff,3)
             )
      )
tCritical <- qt(.05/2,lower.tail=FALSE,df=dfDiff)
print(paste0("The 95% CI for difference in means is: ",
             paste(round(muDiff + c(-1,1) * seDiff * tCritical,3),collapse=" , ")
             )
      )

## Compare with R
t.test(myVar1,myVar2,paired=TRUE)
```

As expected, the results from R match the associated hand calculations.  
  
  
## Module 2: Categorical Association  
Tests for categorical association check relative frequencies for two or more categories.  The typical test statistic is chi-squared.  Comparisons are made against an "expected frequency" which can be a theoretical frequency or the frequency that would be obtained by multiplying out row/column frequencies.  

The general approach to a test matrix is to put the explanatory variables in the rows and the response variable in the columns.  There is, of course, no actual meaningful correlation statistic since the column orders are arbitrary.  So, the general framing would be:

* Ho: The variables are all independent, which is to say P(row, column) = P(row) * P(column)  
* Ha: There are one or more dependencies  
* Approach:  Create a new table (expected) that is P(row) * P(column)  

The example from the UvA course is used a few times in this module:  
```{r}
artHistory <- data.frame(fruit=c(11,8,3), flowers=c(5,6,10),mixed=c(1,8,12),
                         row.names=c("early","late","Baroque")
                         )
## Raw data frame
artHistory

## Joint frequencies
jointFreqArt <- artHistory/sum(artHistory)
jointFreqArt

## Marginal (column) frequencies
margColArt <- colSums(artHistory)/sum(artHistory)
t(as.matrix(margColArt))

## Marginal (row) frequencies
margRowArt <- rowSums(artHistory)/sum(artHistory)
as.matrix(margRowArt)

## Expected frequencies = product of marginal frequencies
expArtHistory <- ( as.matrix(margRowArt) %*% t(as.matrix(margColArt)) ) * sum(artHistory)
round(expArtHistory,1)
```

####_Chi-squared test for association_  
The chi-squared statistic is then the sum over all cells of residual^2 / expected, which could also be formulated as sum over all cells of (observed - expected)^2 / expected.  

The chi-squared statistics can then be assessed for its likelihood:  
* Test-statistic: sum over all cells of residual^2/expected  
* df = (nrow - 1) * (ncol - 1)  
* The mean of a chi-squared distribution with df=n will be n  
* The chi-squared is always greater than or equal to zero  
* pchisq(myStat, df=myDF, lower.tail=FALSE) brings back the odds of seeing >= myStat given myDF  

Note that we require n=5+ for each expected cell for this statistic to be reasonable.  

```{r}
## The residuals data frame
resArtHistory <- artHistory - expArtHistory
resArtHistory

## Test Statistic method 1 -- residual^2 / expected
testMatrix1 <- resArtHistory^2 / expArtHistory
testMatrix1
testStat1 <- sum(testMatrix1)
testStat1

## Test statistics method 2 -- (observed/expected - 1)^2
testMatrix2 <- (artHistory - expArtHistory)^2 / expArtHistory
testMatrix2
testStat2 <- sum(testMatrix2)
testStat2

## Calculate df
dfArt <- (ncol(artHistory) - 1) * (nrow(artHistory) - 1)
dfArt

## Report on pStatistic
pStat <- pchisq(testStat1, df=dfArt, lower.tail=FALSE)
print(paste0("Chi-squared is ",round(testStat1,2)," on df=",dfArt," for p=",round(pStat,4)))

## No surprise that this matches R
chisq.test(artHistory,correct=FALSE)
```

####_Interpretation of chi-squared_  
There are two ways to help with interpreting chi-squared.  Recall that as df increases, you expect chi-squared to increase, so there is no specific meaning to "this has chi-squared 6 while that has chi-squared 12".  

* Cramer's V is an attempt to standardize the chi-squared to within 0-1 where 0 means "no association" and 1 means "perfect association".  The challenge is that V tends to increase no matter what as the matrix becomes less square, so it is still not a perfect test.  
* Standarized residuals give a sense for how far each cell in the observed matrix is from expected.  It is reported on either a Pearson basis or a residuals basis.  

The calculation for Cramer's V is as follows:  
  
* V = sqrt(chi-squared / (n * m))  
* n = total cells in the matrix  
* m = lesser of # rows or # columns, then subtract 1  
  
There are two means of calculating the standardized residuals:  
  
* Pearson - residual / sqrt(expected)  
* Standraized - resiudal / sqrt(V)  
* V for a given cell =  expNCell * (1 - Prow) * (1 - Pcol)  
* The standardized residuals follow a z-distribution  
  
The code chunks below are copied from CheckChiSq_v001.Rmd:  
  
```{r}
testFrame <- data.frame(colI = c(5, 34, 33), colII = c(6, 47, 32), 
                        colIII = c(9, 48, 14), row.names=c("A","B","C")
                        )
testFrame

expMatrix <- as.matrix(rowSums(testFrame)) %*% t(as.matrix(colSums(testFrame)))
expMatrix <- expMatrix / sum(testFrame)
round(expMatrix,1)

resMatrix <- testFrame - expMatrix
round(resMatrix,1)

testChiSq <- sum(resMatrix^2 / expMatrix)
nR <- nrow(resMatrix)
nC <- ncol(resMatrix)

dfChiSq <- (nR - 1) * (nC - 1)
cramerV <- sqrt(testChiSq / (nR * nC) / (min(nR, nC) - 1) )

## Pearson residuals - (obs - exp) / sqrt(exp)
prsMatrix <- resMatrix / sqrt(expMatrix)

## Standardized residuals - (obs - exp) / sqrt(V)
## Vij = sqrt(Expij * (1 - pRow) * (1 - pCol))
stdMatrix <- resMatrix / 
             sqrt(expMatrix * 
                   (as.matrix(1 - rowSums(expMatrix)/sum(expMatrix)) 
                    %*%             
                    t(as.matrix(1-colSums(expMatrix)/sum(expMatrix)))
                    )
                  )

print(paste0("Chi-squared is ",round(testChiSq,2)," with df=",dfChiSq,
             " (p=",
             round(pchisq(testChiSq, df=dfChiSq, lower.tail=FALSE),4),
             ")"
             )
      )

print(paste0("Cramer's V is: ",round(cramerV,3)))

## Pearson residuals - (obs - exp) / sqrt(exp)
round(prsMatrix,2)

## Standardized residuals - (obs - exp) / sqrt(V)
## Vij = sqrt(Expij * (1 - pRow) * (1 - pCol))
round(stdMatrix,2)

testChi <- chisq.test(testFrame)
for (intCtr in 1:length(testChi)) {
    print(testChi[intCtr])
}

## Borrowed from http://www.r-bloggers.com/example-8-39-calculating-cramers-v/
## With adaptations
cv.test <- function(x) {
    CV <- sqrt(chisq.test(x, correct=FALSE)$statistic /
               (sum(!is.na(x)) * (min(nrow(x), ncol(x)) - 1) )
               )
  
    print.noquote("Cramér V / Phi:")
  
    return(as.numeric(CV))
}

cv.test(testFrame)
```

The R functions (chisq.test and cv.test) match with the hand calculations as expected.  

####_Running chi-squared as "goodness of fit"_  
The chi-squared test can also be run to compare some actual observations against a theoretical distribution.  There are a few modest changes:  
  
* Ho: Observed does not differ from expected based on theory  
* Ha: Observed differs in at least some regard from expected based on theory  
* Test statistic chi-squared is sum-over-columns of (observed - expected)^2 / expected  
* df = N-1 where there are N columns explored  
* Requirement that expected be 5+ in each column  

Expected is generally a vector of probabilities summing to 1, with expected for each column become T*expected where T is the sum across all of the observed elements.  

An obvious but important caution is that this is absolutely NOT an appropriate way to compare two rows of observed data against each other!  This only works when you have a theoretical expectation prior to experimentation, and an observed dataset during your experiment plainly does not qualify!  

See below for a very simple example:  
```{r}
myVector <- c(5,20,10,35)
myTheory <- c(.2,.3,.1,.4)

chiSqAssoc <- (myVector - (sum(myVector) * myTheory) )^2 / ( sum(myVector) * myTheory) 
chiSqAssoc
dfAssoc <- length(chiSqAssoc) - 1
pStat <- pchisq(sum(chiSqAssoc), df=dfAssoc, lower.tail=FALSE)

print(paste0("Chi-squared is ", round(sum(chiSqAssoc),2),
             " with df=",dfAssoc," for p=",round(pStat,4)
             )
      )

chisq.test(myVector, correct=FALSE,p=myTheory)
```

As expected, the results from R functions match the hand calculations.  

####_Side notes for chi-squared tests_  
Sometimes, not all of the conditions for chi-squared testing are met.  To wit:  
  
* Data from a non-random sample (frequent error; make sure your experiment is properly randomized)  
* Categories are not exclusive (make sure you have a column for Both-A-and-B and do not count +1 in both the A and B columns)  
* Categories are not exhaustive (make sure everything can be placed somewhere, even if just All Other)  
* Sample too small; requires expected values of 5+ per call (aggregate as needed if not)  
* Interpreting chi-squared as goodness of association (no way - Cramer's V at least gets you in the right direction)  
* Interpreting a high p-value as "support of theory" (as with all hypothesis testing, all we will ever do is fail to reject the null hypothesis; we do not accept/confirm the null!)  
  
####_Fisher's exact test_  
The Fisher's exact test is the solution when you have a contained, small-n problem which precludes using chi-squared tests of association.  
  
The test is generally designed for a 2x2 table as follows:  
  
  
a       | b       | (a+b)      
------- | ------- | -----------  
c       | d       | (c+d)      
(a+c)   | (b+d)   | (a+b+c+d)  
  
  
Define n=(a+b+c+d).  This configuration can occur (a+c)! * (b+d)! * (a+b)! * (c+d)! / [n! * a! * b! * c! * d!] of the time.  So, if a were the critical value, you could test how likely (one-sided or two-sided) you are to get the specific or more extreme value for a.

Recall that if you do not have a small-n problem, you use chi-squared!  This test is something of a mess.  See associated R code:  

```{r}
nTotal <- 23
nAplusB <- 10
nAplusC <- 8

fishStore <- data.frame(a=0:min(nAplusB,nAplusC),probA=rep(-1,min(nAplusB,nAplusC)+1))

for (intCtr in 1:(min(nAplusB,nAplusC)+1)) {
    a <- intCtr - 1
    b <- nAplusB - a
    c <- nAplusC - a
    d <- nTotal - a - b - c
    numer1 <- factorial(a+c) * factorial(b+d)
    numer2 <- factorial(a+b) * factorial(c+d)
    denom1 <- factorial(nTotal)
    denom2 <- factorial(a) * factorial(b) * factorial(c) * factorial(d)
    fishStore[intCtr,"probA"] <- (numer1/denom1) * (numer2/denom2)
}

fishStore
sum(fishStore[c(0,8),"probA"])

fisher.test(matrix(data=c(0,10,8,5),nrow=2,byrow=TRUE))$p.value
fisher.test(matrix(data=c(1,9,7,6),nrow=2,byrow=TRUE))$p.value
fisher.test(matrix(data=c(2,8,6,7),nrow=2,byrow=TRUE))$p.value
fisher.test(matrix(data=c(3,7,5,8),nrow=2,byrow=TRUE))$p.value
fisher.test(matrix(data=c(4,6,4,9),nrow=2,byrow=TRUE))$p.value
fisher.test(matrix(data=c(5,5,3,10),nrow=2,byrow=TRUE))$p.value
fisher.test(matrix(data=c(6,4,2,11),nrow=2,byrow=TRUE))$p.value
fisher.test(matrix(data=c(7,3,1,12),nrow=2,byrow=TRUE))$p.value
fisher.test(matrix(data=c(8,2,0,13),nrow=2,byrow=TRUE))$p.value

```

These still do not match perfectly.  I get the sense this is a rarely used test, but time permitting it may be good to figure out the discrepancy.  
  
## Module 3: Simple Regression  